---
title: "EAB spatial analysis in R"
output: html_document
date: "2023-01-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ape)
library(geoR)
```

## Data

I started with the 20 EAB and 20 non-EAB data sets within the larger EAB_PROP_DATA_2.xlsx file. If we want to change the cutoff to a higher number, we can use an exclude function in R to chop out rows with <40 or <50 records.


```{r input}


eab_dat <- read.csv("./Data/Originals/20 EAB data.csv")

noneab_dat <- read.csv("./Data/Originals/20 nonEAB.csv")

#rename columns
cnames <- c("Town", "Y2012", "Y2013", "Y2014", "Y2015", "Y2016", "Y2017", "Y2018", "Y2019", "Y2020", "Y2021", "Y2022", "Totals")

colnames(eab_dat) <- cnames
colnames(noneab_dat) <- cnames

#import lat and long from EAB_PROP_DATA.xslx. Converted to csv in excel to avoid import issues

waypoints_dat <- read.csv("./Data/Originals/EAB_PROP_DATA.csv") %>% 
  dplyr::select(c("Town","Lat","Long")) %>% # we only need the gps for towns from this dataset
  rename(lon = Long)

# merge waypoints onto both, drop the OG
eab_dat <- eab_dat %>% left_join(waypoints_dat, by="Town") 

noneab_dat <- noneab_dat %>% left_join(waypoints_dat, by="Town")
   
```

## Summary of Data

Visual inspection of the first several rows to make sure we're all set

```{r data structure}
head(eab_dat)

```

## Moran's I

# Here I am testing to see if we can get some morans I tests for spatial autocorrelation
# No p-value estimates for Y2012, but a full run with no errors on Y2021

```{r morans}
# https://stats.oarc.ucla.edu/r/faq/how-can-i-calculate-morans-i-in-r/#:~:text=Moran's%20I%20is%20a%20measure,and%20then%20library(ape).


eab_dists <- as.matrix(dist(cbind(eab_dat$Lon, eab_dat$Lat)))

# inverse and replace diagonals with zeros to prepare matrix for Moran.I

eab_dists_inv <- 1/eab_dists
diag(eab_dists_inv) <- 0

# morans test for 2012
Moran.I(eab_dat$Y2012, eab_dists_inv, na.rm = TRUE)

# morans test for 2021
Moran.I(eab_dat$Y2021, eab_dists_inv, na.rm = TRUE)

```

## First go at krieging

```{r krieging prep}
# https://stats.oarc.ucla.edu/r/faq/how-do-i-generate-a-variogram-for-spatial-data-in-r/



summary(eab_dists)

```
