---
title: "EAB spatial analysis in R"
output: html_document
date: "2023-01-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ape)
library(gstat)
```

## Data

I started with the 20 EAB and 20 non-EAB data sets within the larger EAB_PROP_DATA_2.xlsx file. If we want to change the cutoff to a higher number, we can use an exclude function in R to chop out rows with <40 or <50 records.


```{r input}


eab_dat <- read.csv("./Data/Originals/20 EAB data.csv")

noneab_dat <- read.csv("./Data/Originals/20 nonEAB.csv")

#rename columns
cnames <- c("Town", "Y2012", "Y2013", "Y2014", "Y2015", "Y2016", "Y2017", "Y2018", "Y2019", "Y2020", "Y2021", "Y2022", "Totals")

colnames(eab_dat) <- cnames
colnames(noneab_dat) <- cnames

#import lat and long from EAB_PROP_DATA.xslx. Converted to csv in excel to avoid import issues

waypoints_dat <- read.csv("./Data/Originals/EAB_PROP_DATA.csv") %>% 
  dplyr::select(c("Town","Lat","Long")) %>% # we only need the gps for towns from this dataset
  rename(lon = Long, lat = Lat) # quibble with upper cases

# merge waypoints onto both, drop the OG
eab_dat <- eab_dat %>% left_join(waypoints_dat, by="Town") 

noneab_dat <- noneab_dat %>% left_join(waypoints_dat, by="Town")
   
```

## Summary of Data and Proportions

Visual inspection of the first several rows to make sure we're all set

```{r data structure}
head(eab_dat)
head(noneab_dat)

dataframe_proportions <- eab_dat

#loop through the columns
for(i in 3:12){
    #calculate the proportion for the current column
    dataframe_proportions[,i] <- ifelse(noneab_dat[,i] == 0, NA, eab_dat[,i]/noneab_dat[,i])
}


```

## Moran's I

# Here I am testing to see if we can get some morans I tests for spatial autocorrelation
# No p-value estimates for Y2012, but a full run with no errors on Y2021

```{r morans}
# https://stats.oarc.ucla.edu/r/faq/how-can-i-calculate-morans-i-in-r/#:~:text=Moran's%20I%20is%20a%20measure,and%20then%20library(ape).

# Morans I for all EAB only is a mix of sig and non sig
# Morans I for all beetles total is all high p-values (>.6)
# Morans I for proportions is highly autocorrleated with p-values all < 0.001

# using dataframe_proportions following arcgis models and R results

eab_dists <- as.matrix(dist(cbind(dataframe_proportions$lon, dataframe_proportions$lat)))

# inverse and replace diagonals with zeros to prepare matrix for Moran.I
eab_dists_inv <- 1/eab_dists
diag(eab_dists_inv) <- 0

# morans test for 2021 as a test case with abundance only EAB data
Moran.I(dataframe_proportions$Y2021, eab_dists_inv, na.rm = TRUE)

# loop to run Moran's I across all 10 years
results <- list()
for (i in 2012:2022) {
  var_name <- paste0("Y", i)
  command <- paste0("Moran.I(dataframe_proportions$", var_name, ", eab_dists_inv, na.rm = TRUE)")
  result <- eval(parse(text = command))
  results[[i-2011]] <- result #this makes the loop start after 2011
}

# reformat the output from the loop above into a legible table
results_df <- data.frame(year = paste0("Y", 2012:2022),
                        observed = sapply(results, function(x) x$observed),
                        expected = sapply(results, function(x) x$expected),
                        sd = sapply(results, function(x) x$sd),
                        p.value = sapply(results, function(x) x$p.value))

results_df

```

## First go at kriging

```{r variograms}
# https://zia207.github.io/geospatial-r-github.io/semivariogram-modeling.html


```


## First go at variograms

```{r variograms}
# https://zia207.github.io/geospatial-r-github.io/semivariogram-modeling.html


```
